Sequoidal Categories and Transfinite Games: Towards a Coalgebraic Approach to Linear Logic\footnote{This work was partially supported by [TODO: find the appropriate grant codes]}

### 2017-04-09 round-corner/paper/lipics-v2016-authors/sequoidal.tex
Changed direction of composition (...)
Rewrote introduction.

%% Other macros
\newcommand*\from{\colon}
\def \inv {^{-1}}
\DeclareMathOperator{\id}{id}
\DeclareMathOperator{\pr}{pr}
\newcommand{\tensor}{\otimes}
\newcommand{\sequoid}{\oslash}
\renewcommand{\implies}{\multimap}
\newcommand{\comp}[2]{#1 \circ #2}
\newcommand{\cprd}{\sqcup}
\newcommand{\bigcprd}{\bigsqcup}
\newcommand{\C}{\mathcal C}
\newcommand{\F}{\mathcal F}
\newcommand{\G}{\mathcal G}
\newcommand{\W}{\mathcal W}
\newcommand{\suchthat}{\;\colon\;}
\newcommand{\varsuchthat}{\;\mid\;}
\newcommand{\OP}{\{O,P\}}
\newcommand{\emptyplay}{\epsilon}
\newcommand{\prefix}{\sqsubseteq}
\newcommand{\pprefix}{\sqsubsetneqq}
\newcommand{\ppprefix}{\sqsubset}
\DeclareMathOperator{\assoc}{assoc}
\DeclareMathOperator{\lunit}{lunit}
\DeclareMathOperator{\runit}{runit}
\DeclareMathOperator{\sym}{sym}
\newcommand{\braid}{\sym}
\newcommand{\blank}{\;\underline{\hspace{1.5ex}}\;}
\newcommand{\der}{{\mathtt{der}}}
\newcommand{\mult}{{\mathtt{mult}}}
\DeclareMathOperator{\wk}{wk}
\newcommand{\toisom}{{\xrightarrow{\cong}}}
\newcommand{\passoc}{{\mathtt{passoc}}}
\newcommand{\pcomm}{{\mathtt{pcomm}}}
\newcommand{\run}{{\mathtt{r}}}
\newcommand{\lun}{{\mathtt{l}}}
\newcommand{\fcoal}[1]{{\leftmoon #1 \rightmoon}}
\renewcommand{\subset}{\subseteq}
\let\oldemptyset\emptyset
\let\emptyset\varnothing
\newcommand{\s}{\mathfrak{s}}
\newcommand{\dist}{{\mathtt{dist}}}
\newcommand{\dec}{{\mathtt{dec}}}
\renewcommand{\int}{{\mathtt{coh}}}
\renewcommand{\coh}{\int}
\DeclareMathOperator{\CCom}{CCom}
\newcommand{\catname}[1]{{\mathsf{#1}}}
\newcommand{\Set}{\catname{Set}}
\newcommand{\Ord}{\catname{Ord}}
\newcommand\oppcat[1]{#1^{\mathrm{op}}}
\newcommand{\bN}{{\mathbb{N}}}
\newcommand{\OK}{{\mathtt{OK}}}
\newcommand{\zero}{{\mathtt{zero}}}
\renewcommand{\succ}{{\mathtt{succ}}}
\newcommand{\Var}{{\mathtt{Var}}}
\renewcommand{\read}{{\mathtt{read}}}
\newcommand{\wwrite}{{\mathtt{write}}}
\newcommand{\Rel}{\catname{Rel}}
\newcommand{\inr}{{\mathsf{inr}}}
\newcommand{\inl}{{\mathsf{inl}}}
\newcommand{{\Na}}{\bN}
\newcommand{{\cell}}{{\mathsf{cell}}}
\newcommand{\fix}{{\mathsf{fix}}}
\newcommand{\eq}{{\mathsf{eq}}}
\newcommand{\co}{{\textrm{\textexclamdown}}}

@@

Games models of pure functional programming languages such as PCF (\cite{ajmPcf} and \cite{hoPcf}, for example) tend to include the requirement that strategies be \emph{history-free}: that is, the response to a particular position may depend only upon the last move played, and not upon what has gone before.  This makes sense for stateless languages, where the program should have no `memory' to recall what has gone before.

If we include history-sensitive strategies in our models, then we can start to recover some stateful behaviour (see \cite{SamsonGuyIAPassive}, for example), since history-sensitive strategies have a built in state.  This concept of state is often hard to work with: since the representation as a strategy is so intensional, there is no object representing the state, and the state of the program must be inferred from the behaviour of the strategy.

At the other extreme, we have traditional, domain-theoretic semantics, in which stateful programs are represented as \emph{state transformers} -- that is, functions $W\times X \to W\times Y$, where $W$ is an object representing the state.

State transformers are typically fairly straightforward to reason about, and we would like to have a similar technique available in our games models.  This gap was bridged by one of the authors in \cite{laird02} with the introduction of the new \emph{sequoid} operator $\blank\sequoid\blank$ on games, which the same as the tensor product $\blank\tensor\blank$, except that play is required to start in the left hand game.  We interpret a state transformer as a coalgebra for the functor $A\sequoid\blank$ -- that is, a morphism from $S$ to $A\sequoid S$, so that the state $S$ is only modified after the program $A$ has been consulted.  The benefit of this explicit ordering is that we may model multiple iterations of the state transformer as coalgebras:
\[
  S \to A \sequoid (A \sequoid \cdots \sequoid (A \sequoid S)\cdots)
  \]
where the copies of $A$ are started in order from left to right.  The state from a previous invocation of $A$ is passed on to the next, until the final modified state finally appears at the right.

Now we introduce the key part of the construction: suppose that the functor $A\sequoid\blank$ has a final coalgebra $\oc A\to A\sequoid \oc A$.  Then a coalgebra $S\to A\sequoid S$ (i.e., a state transformer) naturally gives rise to a morphism $S \to \oc A$ in which the state is completely encapsulated and we are allowed to invoke the program $A$ as many times as we like.  We will show that we can construct various history-sensitive strategies denoting stateful programs using precisely this strategy; in particular, we will give a coalgebraic construction of the $\cell$ strategy from \cite{SamsonGuyIAPassive}, which is the key to their model for Idealized Algol.

Even more intriguingly, the game $\oc A$ that arises as the final coalgebra for the functor $A\sequoid\blank$ has the structure of an exponential object in the category of games: it is the cofree commutative comonoid over $A$.  We will spend the rest of the paper examining why this is, and what general conditions we need in order to show that the final coalgebra for $\oc A$ and the cofree commutative comonoid over $A$ coincide.  For this, we will need the abstract definition of a \emph{sequoidal category}, which is an extension of a monoidal category in order to accommodate the sequoid operator.  We will first consider the explicit formula for the cofree commutative comonoid given by Melli\'es, Tabareau and Tasson in \cite{MelliesCofCommCom} and show that it can be given a sequoidal interpretation.  The Melli\'es-Tabareau-Tasson construction works for finite games, but its hypotheses do not hold in the case of the category of games with infinitary winning conditions and winning strategies, even though the final coalgebra for $A\sequoid\blank$ and the cofree commutative comonoid over $A$ still coincide in these categories.

Instead, we will show that if our sequoidal category has all products then the cofree commutative comonoid structure on $\oc A$ may be constructed directly from the sequoidal structure and the definition of $A\sequoid\blank$ as a final coalgebra for $A\sequoid\blank$, provided that we insist on a further hypothesis: that the functor $\oc$ is monoidal from the Cartesian category to the monoidal category: i.e., if we have an isomorphism between $\oc A \tensor \oc B$ and $\oc(A\times B)$ for all $A,B$.  We show that this hypothesis is automatically satisfied whenever $\oc A$ has the structure of a \emph{bifree algebra} for $A\sequoid\blank$; i.e., when it simultaneously has the structure of final coalgebra and initial algebra.  However, the exponential in the category of win-games is not an initial algebra for $A\sequoid\blank$ \cite{martinsthesis}.

Lastly, we will provide an illuminating counterexample: a category of games in which the final coalgebra for $A\sequoid\blank$ and the cofree commutative comonoid over $A$ do not coincide.  This is the category of games where plays are allowed to have transfinite length.  We will show that the Melli\'es-Tabareau-Tasson construction fails badly in this setting, but that we may still construct the final coalgebra for $A\sequoid\blank$ as the limit of its final sequence.  In this setting, the final coalgebra is the game where we have `only' $\omega$-many copies of the sub-game to play in, whereas in the cofree commutative comonoid we may open transfinite sequences of the subgame.

@@

Our sequoidal category will be closely modelled upon the category of games we have just considered: the objects will be games, with the modification that sequences of moves may now have transfinite length.  This is a natural construction, occurring in the study of determinacy by Mycielski \cite{mycielskiChoice}, Blass \cite{blassChoice} and Weiss \cite{longgamesbook}, and it appears to be present in the semantic context in the work of Roscoe \cite{RoscoeCspInfinite}, Levy \cite{LevyGsInfinite} and Laird \cite{LairdOrdinalGames}.

>>>>

%% Other macros
\newcommand*\from{\colon}
\def \inv {^{-1}}
\DeclareMathOperator{\id}{id}
\DeclareMathOperator{\pr}{pr}
\newcommand{\tensor}{\otimes}
\newcommand{\sequoid}{\oslash}
\renewcommand{\implies}{\multimap}
\newcommand{\comp}[2]{#2 ; #1}
\newcommand{\cprd}{\sqcup}
\newcommand{\bigcprd}{\bigsqcup}
\newcommand{\C}{\mathcal C}
\newcommand{\F}{\mathcal F}
\newcommand{\G}{\mathcal G}
\newcommand{\W}{\mathcal W}
\newcommand{\suchthat}{\;\colon\;}
\newcommand{\varsuchthat}{\;\mid\;}
\newcommand{\OP}{\{O,P\}}
\newcommand{\emptyplay}{\epsilon}
\newcommand{\prefix}{\sqsubseteq}
\newcommand{\pprefix}{\sqsubsetneqq}
\newcommand{\ppprefix}{\sqsubset}
\DeclareMathOperator{\assoc}{assoc}
\DeclareMathOperator{\lunit}{lunit}
\DeclareMathOperator{\runit}{runit}
\DeclareMathOperator{\sym}{sym}
\newcommand{\braid}{\sym}
\newcommand{\blank}{\;\underline{\hspace{1.5ex}}\;}
\newcommand{\der}{{\mathtt{der}}}
\newcommand{\mult}{{\mathtt{mult}}}
\DeclareMathOperator{\wk}{wk}
\newcommand{\toisom}{{\xrightarrow{\cong}}}
\newcommand{\passoc}{{\mathtt{passoc}}}
\newcommand{\pcomm}{{\mathtt{pcomm}}}
\newcommand{\run}{{\mathtt{r}}}
\newcommand{\lun}{{\mathtt{l}}}
\newcommand{\fcoal}[1]{{\leftmoon #1 \rightmoon}}
\renewcommand{\subset}{\subseteq}
\let\oldemptyset\emptyset
\let\emptyset\varnothing
\newcommand{\s}{\mathfrak{s}}
\newcommand{\dist}{{\mathtt{dist}}}
\newcommand{\dec}{{\mathtt{dec}}}
\renewcommand{\int}{{\mathtt{coh}}}
\renewcommand{\coh}{\int}
\DeclareMathOperator{\CCom}{CCom}
\newcommand{\catname}[1]{{\mathsf{#1}}}
\newcommand{\Set}{\catname{Set}}
\newcommand{\Ord}{\catname{Ord}}
\newcommand\oppcat[1]{#1^{\mathrm{op}}}
\newcommand{\bN}{{\mathbb{N}}}
\newcommand{\OK}{{\mathtt{OK}}}
\newcommand{\zero}{{\mathtt{zero}}}
\renewcommand{\succ}{{\mathtt{succ}}}
\newcommand{\Var}{{\mathtt{Var}}}
\renewcommand{\read}{{\mathtt{read}}}
\newcommand{\wwrite}{{\mathtt{write}}}
\newcommand{\Rel}{\catname{Rel}}
\newcommand{\inr}{{\mathsf{inr}}}
\newcommand{\inl}{{\mathsf{inl}}}
\newcommand{{\Na}}{\bN}
\newcommand{{\cell}}{{\mathsf{cell}}}
\newcommand{\fix}{{\mathsf{fix}}}
\newcommand{\eq}{{\mathsf{eq}}}
\newcommand{\co}{{\textrm{\textexclamdown}}}

@@

Games models of pure functional programming languages such as PCF (\cite{ajmPcf} and \cite{hoPcf}, for example) tend to include the requirement that strategies be \emph{history-free}\footnote{In fact, the model in \cite{hoPcf} and other models following the `Hyland-Ong' tradition use \emph{innocent} strategies, rather than history free strategies to model statelessness.  We will focus on AJM-style Game Semantics in this paper, so we shall focus on history-freeness vs. history sensitivity rather than innocence.}: that is, the response to a particular position may depend only upon the last move played, and not upon what has gone before.  This makes sense for stateless languages, where the program should have no `memory' to recall what has gone before.

If history-free strategies are the game-semantic interpretation of stateless programs, then history-sensitive strategies should be the game-semantic interpretation of stateful programs, and this is indeed how they are used: Abramsky and McCusker's model of Idealized Algol \cite{SamsonGuyIAPassive}, for example, makes use of history-sensitive strategies in order to model stateful programs.  However, this idea of state can be difficult to work with, because it is completely \emph{implicit} in the strategy.  Let us, for example, try to model a simple storage cell.  Let $X$ be a set of values and, as in \cite{SamsonGuyIAPassive}, let $\Var[X]$ be the game with maximal plays $rx$ and $w_x\checkmark$, where $x$ ranges over $X$.  $rx$ corresponds to a $\read$ method: player $O$ plays $r$ and expects to receive the value $x$ currently in the cell, while $w_x\checkmark$ corresponds to a write method: player $O$ writes $x$ into the cell and then player $P$ confirms that the write has taken place.

It is now clear that we have a strategy $\cell$ on $\oc \Var[X]$ corresponding to a storage cell: whenever player $O$ plays $w_x$, respond with $\checkmark$, and whenever player $O$ plays $r$, respond with the last $x$ such that $w_x$ was played (if the cell is uninitialized -- i.e., if $w_x$ has never been played for any $x$ -- then we return either nothing or some specified initial value, depending on what we are trying to model).  This strategy is clearly history sensitive and represents a stateful program; however, it is tricky to reason about the statefulness of $\cell$, because there is no object explicitly representing the state: the state in $\cell$ is \emph{implicit}, and it appears that we need to resort to combinatorial arguments in order to reason about it.

Contrast this with the handling of state in traditional domain-theoretic semantics, where we represent a stateful program as a function $W \times X \to W \times Y$, or a \emph{state transformer}, where $W$ is a built-in store of state that can be reasoned about directly.  In 2002, one of the authors introduced the game-semantic equivalent of a state-transformer by defining a new operation on games -- the \emph{sequoid} $\blank\sequoid\blank$\cite{laird02}.  The game $A\sequoid B$ is the same as the game $A\tensor B$, except that the first move must take place in $A$.  We now model a state transformer for a program as a morphism $S \to A \sequoid S$ -- i.e., a \emph{coalgebra} for the functor $A\sequoid\blank$ -- where $A$ is the program and $S$ is the changed state.  The justification for the order relation on the sequoid operator is that the state should only change in response to some action by the environment.

For example, let $X$ be the game with maximal plays $rx$ for $x\in X$.  Then we may define a state transformer for $X \implies \Var[X]\sequoid X$ as follows: play must start in the game $\Var[X]$.  If player $O$ starts with the move $r$, then request a value of $x$ on the left and then copy it into both $\Var[X]$ and $X$ on the right (so if $X$ on the left represents the initial state and $X$ on the right represents the final state, then we have returned the value of the state and left the state unchanged).  If player $O$ starts with the move $w_y$, return $\checkmark$ and then copy the value $y$ over to $X$ on the right, ignoring the value of $X$ on the left.

Note that we may model multiple iterations of the state transformer as coalgebras:
\[
  S \to A \sequoid (A \sequoid \cdots \sequoid (A \sequoid S)\cdots)
  \]
where the copies of $A$ are started in order from left to right.  The state from a previous invocation of $A$ is passed on to the next, until the final modified state finally appears at the right.

Now we come to the key point of the construction: if $A$ is a game, then the Hyland exponential $\oc A$ \cite{hyland1997games} has the structure of the \emph{final coalgebra} for the functor $A \sequoid\blank$.  Therefore, following \cite{jacobs}, any coalgebra $S\to A\sequoid S$ (i.e., any state transformer) gives rise to a morphism $S\to\oc A$ that encapsulates the explicit state of the state transformer into an implicitly stateful history-sensitive strategy $S\to\oc A$ that corresponds to repeatedly invoking the state transformer, passing the state from one invocation to the next.

Indeed, the state transformer we defined on $X\implies\Var[X]\sequoid X$ gives rise to a morphism $X\to\oc\Var[X]$ that gives us the behaviour of a storage cell with initial value given by the copy of $X$ on the left; composing this with our desired initial value on $X$ gives us the strategy $\cell$.  The difference now is that, rather than reason directly about the implicit strategy $\cell$, we may reason about the explicit state in our state transformer and then use coalgebraic methods to talk about $\cell$.

It is clear that the key fact in the construction is that the final coalgebra for $A\sequoid\blank$ coincides with the exponential $\oc A$, which has the structure of the \emph{cofree commutative comonoid} on $A$.  We will begin by introducing the general idea of using coalgebra to define higher-order stateful objects, using a construction on the category $\Rel$ of sets and relations.  This construction is simple to define, but the final coalgebra in that context does not give us a good model of the exponential: it gives us a cofree comonoid, but not a commutative one.  This motivates us to study state transformers using game semantics and the sequoid operator.

We review the definition of a \emph{sequoidal category} from \cite{laird02}: a category-theoretic formalization of the sequoid operator akin to the formalization of the tensor product $\tensor$ by the definition of a \emph{monoidal category}.  We illustrate the definition by showing how the sequoidal structure allows us to model stateful programs as morphisms into the final coalgebra for the functor $A\sequoid\blank$, using the example of the $\cell$ strategy in the category of games.

We spend the rest of the paper discussing the question of what assumptions we need in order to prove that the final coalgebra for $A\sequoid\blank$ coincides with the cofree commutative comonoid on $A$.  The explicit formula for the cofree commutative comonoid given by Melli\'es, Tabareau and Tasson \cite{MelliesCofCommCom} provides a way to link the two constructions that works for the basic category of games; however, it is not powerful enough to prove that the final coalgebra and the cofree commutative comonoid are the same in the category of games with infinitary winning conditions.

We turn instead to a purely sequoidal/coalgebraic proof that the final coalgebra $\oc A$ for $A\sequoid\blank$ is the carrier of the cofree commutative comonoid on $A$.  In this case, we will need to assume that our category has a Cartesian structure and also assume an additional hypothesis: that if $A$ and $B$ are objects, then a certain natural morphism $\oc A \tensor \oc B \to \oc (A\times B)$ is an isomorphism.  We will refer to this as the \emph{strong monoidal hypothesis}, since it can be used to show that $\oc\blank$ gives rise to a strong monoidal functor from the category with its Cartesian structure to the category with its monoidal functor.

We show that the strong monoidal hypothesis holds whenever $\oc A$ is a \emph{bifree algebra} for $A\sequoid\blank$: that is, it is simultaneously an initial algebra and a final coalgebra.

We finish with a non-example: the sequoidal category of games with transfinite plays.  In this setting the final coalgebra for $A\sequoid\blank$ does not carry the structure of the cofree commutative comonoid over $A$.  By our discussion above, note that the final coalgebra for $A\sequoid\blank$ has the structure of $\omega$-many copies of the game $A$, where we pass the state from $A_i$ to $A_{i+1}$:
\tikzset{dotsstyle/.style={draw=none}}
\[
  \begin{tikzpicture}
    \matrix[matrix of nodes, column sep=36pt, inner xsep=1ex, name=m, nodes={draw=black}] {
      $A_0$ & $A_1$ & $A_2$ & |[dotsstyle]| $\cdots$ \\ };
    \path[commutative diagrams/.cd, every arrow, every label, labels={font size=tiny}]
      (m-1-1) edge node {state} (m-1-2)
      (m-1-2) edge node {state} (m-1-3)
      (m-1-3) edge node {state} (m-1-4);
  \end{tikzpicture}
\]

For finite interactions, we can use this to model the exponential, in which repeated access to the resource $A$ is allowed, by passing an empty state in between successive invocations of $A$.  In the transfinite case, however, there might not be enough copies of $A$ in the final coalgebra to satisfy our needs: consider a program that, over time $\omega$, uses the resource $A$ infinitely many times.  If we carry on into transfinite time, the program may continue to request access to the resource $A$.  But this is no longer possible if we are using the final coalgebra for our `exponential', since we are only able to construct a sequence of invocations of $A$ of length $\omega$: there is no predecessor copy of $A$ to receive the (empty) state from.

We shall demonstrate why our earlier approaches fail in this context: the Melli\'es-Tabareau-Tasson construction fails badly in this case, just as it fails for win games and winning strategies (which may be viewed as games of length $\omega+1$).  Similarly, the strong monoidal hypothesis does not hold, since a transfinite interleaving of two sequences of length $\omega$ may have length greater than $\omega$.

@@

Our sequoidal category will be closely modelled upon the category of games we have just considered: the objects will be games, with the modification that sequences of moves may now have transfinite length.  This is a natural construction, occurring in the study of determinacy by Mycielski \cite{mycielskiChoice}, Blass \cite{blassChoice} and Weiss \cite{longgamesbook}.  Transfinite games were used by Berardi and de'Liguoro to give a charactarization of total functionsals \cite{BerardiTransfiniteGames}, and they appear to be present in the semantic context in the work of Roscoe \cite{RoscoeCspInfinite}, Levy \cite{LevyGsInfinite} and Laird \cite{LairdOrdinalGames}.
